Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations

### 2016-02-20 doc/manuscript.tex
clean up introduction with RK changes

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[a4paper]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Thomas McColgan, Hermann Wagner, Richard Kempter},
            pdftitle={Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{Extracellular potentials generated by axonal projections are shaped by
patterns of bifurcations and terminations}
\author{Thomas McColgan, Hermann Wagner, Richard Kempter}
\date{\today}
\usepackage{setspace}
\doublespacing

@@

Extracellular field potentials (EFPs) are at the heart of many
experimental methods used to examine the inner workings of the brain.
This includes invasive (Local Field Potentials, Current Source Density,
Multiunit Activity) as well as noninvasive (EEG, ECoG, ABR) methods
(Nunez and Srinivasan, 2006; Brette and Destexhe, 2012). The origins of
these EFPs, especially in cases in which the activity is not clearly
attributable to a single cell, is a matter of debate.

EFPs in the brain were long thought to be primarily synaptic in origin
(Buzsáki et al., 2012). As a consequence, many modeling studies focus on
the extracellular fields induced by synaptic currents on the dendrites
and soma of the postsynaptic neuron (Holt and Koch, 1999; Gold et al.,
2006; Lindén et al., 2010, 2011; Einevoll et al., 2013; Fernández-Ruiz
et al., 2013). However, a number of recent data analysis and modeling
efforts have revealed that active, non-synaptic membrane currents can
play an important role in generating EFPs (Ray and Maunsell, 2011;
Belluscio et al., 2012; Schomburg et al., 2012; Reimann et al., 2013;
Anastassiou et al., 2015), including far reaching potentials detectable
at the scalp (Teleńczuk et al., 2011, 2015).

One of the reasons for the traditional assumption that axonal currents
contribute little to the EFP is that the far field of an action
potential traveling along an idealized straight axon is quadrupolar,
meaning that its far field decays faster with distance than synaptic
sources which are thought to be dipolar (Nunez and Srinivasan, 2006).
However, in experimental studies that recorded the EFP of axonal
responses, the EFP often has a dipolar structure. For example, Blot and
Barbour (2014) report an EFP with the characteristic dipolar structure
recorded from the vicinity of cerebellar Purkinje cell axons.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  find more examples in literature
\end{itemize}

This discrepancy raises the open question of how axons are able to
generate dipolar field potentials in the actual organism, when
conventional simulation results lead one to expect quardupolar
structures.

The aim of this study is to show how these effects in the EFP of axons
can be explained by the axons' anatomical structure. In particular, we
aim to explain how typical projection patterns in which axons bifurcate
and then terminate in their projection area lead to a dipolar EFP
structure. Such axon bundles, sometimes called nerves or facicles, exist
throughout the peripheral and central nervous system of most animals
(Nornes and Das, 1972; Goodman et al., 1984; Hentschel and Ooyen, 1999;
Kandel et al., 2000). The white matter of the mammalian brain can be
viewed as an agglomeration of such bundles (Schüz and Braitenberg,
2002).

These properties are first investigated by numerical model based on a
forward simulation of the extracellular field potential (Holt and Koch,
1999; Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we show the mechanisms underlying the observed effects by means
of a second, analytically tractable, model of a generic axon bundle.

@@

\emph{Axonal projections generate a dipole-like field potential.} To
understand how axons can generate a dipolar EFP, we simulated a parallel
fiber bundle that at first runs at constant diameter and without
bifurcations, then reaches a bifurcation zone with an increasing number
of fibers, and finally reaches a termination zone, in which the number
of fibers decreases to zero. (Figure 1A).

The spiking activity of a generic axon bundle was simulated by
stimulating the fibers with a background spontaneous firing rate of 100
spikes/second, upon which a pulse of increased activity was added. This
pulse of activity could relate to any kind of evoked activity in the
nervous system, such as a sensory stimulation, motor activity or a
spontaneous transient increase in population spiking activity. We chose
a gaussian pulse with an amplitude of 2900 spikes/second and a standard
deviation of 2.8ms.

To characterize the spatiotemporal dynamics of the evoked EFP, the
potential was calculated for several locations along the axon trunk.
Because we first want to study the effect of the rate pulse on the EFP,
we low-pass filtered the response the responses with a cut-off frequency
of 1kHz. Figures 1A-C all show these low-pass filtered responses. The
distribution of maximum amplitudes of these responses are shown by the
contour lines in Figure 1A. Surrounding the projection zone of the axon
bundle, they show the double-lobed shape typical of a dipole.

Shown alongside the contour lines are dots representing recording
locations, also colored according to the maximum amplitude at their
location. The waveforms recorded at these locations are shown in Figure
1B. The scale of the waveforms are indicated by the color of the trace.
The responses show a biphasic deflection that is elicited by the
population firing rate pulse. This deflection was visible at all
recording locations. It reverses polarity in the middle of the
bifurcation zone of the bundle (red box in Figure 1B). The polarity
reversal occurs by a decrease of the amplitude to zero and subsequent
reappearance with reversed polarity, as opposed to a polarity reversal
through a gradual shift in phase. This behaviour is also typical of a
dipolar field potential.

The point of the polarity reversal coincides roughly with the middle of
the projection zone. This means that the absolute value of the response
amplitude reaches a local minimum, just when the number of axonal fibers
reaches a maximum. To better understand how anatomical features and
response amplitude are related, we compared the value of the EFP
corresponding to the maximum absolute amplitude with the difference in
branchings and terminations (Figure 1C). The amplitude of the maximum
response (red trace) follows the distribution of branchings and
terminations(black histogram).

@@

\hyperdef{}{ref-Anastassiou2015Cell}{}
Anastassiou CA, Perin R, Buzsáki G, Markram H, Koch C (2015) Cell type-
and activity-dependent extracellular correlates of intracellular
spiking. Journal of Neurophysiology 114:608--623 Available at:
\url{http://dx.doi.org/10.1152/jn.00628.2014}.

\hyperdef{}{ref-Belluscio2012CrossFrequency}{}
Belluscio MA, Mizuseki K, Schmidt R, Kempter R, Buzsáki G (2012)
Cross-Frequency Phase--Phase Coupling between Theta and Gamma
Oscillations in the Hippocampus. The Journal of Neuroscience 32:423--435
Available at: \url{http://dx.doi.org/10.1523/jneurosci.4122-11.2012}.

\hyperdef{}{ref-Blot2014Ultrarapid}{}
Blot A, Barbour B (2014) Ultra-rapid axon-axon ephaptic inhibition of
cerebellar Purkinje cells by the pinceau. Nat Neurosci 17:289--295
Available at: \url{http://dx.doi.org/10.1038/nn.3624}.

\hyperdef{}{ref-Brette2012Handbook}{}
Brette R, Destexhe A eds. (2012) Handbook of Neural Activity
Measurement. Cambridge: Cambridge University Press. Available at:
\url{http://dx.doi.org/10.1017/cbo9780511979958}.

\hyperdef{}{ref-Buzsaki2012Origin}{}
Buzsáki G, Anastassiou CA, Koch C (2012) The origin of extracellular
fields and currents--EEG, ECoG, LFP and spikes. Nature reviews
Neuroscience 13:407--420 Available at:
\url{http://dx.doi.org/10.1038/nrn3241}.

\hyperdef{}{ref-carr90}{}
Carr CE, Konishi M (1990) A circuit for detection of interaural time
differences in the brain stem of the barn owl. The Journal of
Neuroscience 10:3227--3246 Available at:
\url{http://www.jneurosci.org/content/10/10/3227.abstract}.

\hyperdef{}{ref-Einevoll2013Modelling}{}
Einevoll GT, Kayser C, Logothetis NK, Panzeri S (2013) Modelling and
analysis of local field potentials for studying the function of cortical
circuits. Nature reviews Neuroscience 14:770--785 Available at:
\url{http://dx.doi.org/10.1038/nrn3599}.

\hyperdef{}{ref-FernandezRuiz2013Cytoarchitectonic}{}
Fernández-Ruiz A, Muñoz S, Sancho M, Makarova J, Makarov VA, Herreras O
(2013) Cytoarchitectonic and Dynamic Origins of Giant Positive Local
Field Potentials in the Dentate Gyrus. The Journal of Neuroscience
33:15518--15532 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0338-13.2013}.

\hyperdef{}{ref-Gold2006Origin}{}
Gold C, Henze DA, Koch C, Buzsáki G (2006) On the origin of the
extracellular action potential waveform: A modeling study. Journal of
neurophysiology 95:3113--3128 Available at:
\url{http://dx.doi.org/10.1152/jn.00979.2005}.

\hyperdef{}{ref-Goldwyn2014Model}{}
Goldwyn JH, Laughlin MM, Verschooten E, Joris PX, Rinzel J (2014) A
Model of the Medial Superior Olive Explains Spatiotemporal Features of
Local Field Potentials. The Journal of Neuroscience 34:11705--11722
Available at: \url{http://dx.doi.org/10.1523/jneurosci.0175-14.2014}.

\hyperdef{}{ref-Goodman1984Cell}{}
Goodman CS, Bastiani MJ, Doe CQ, Lac S du, Helfand SL, Kuwada JY, Thomas
JB (1984) Cell recognition during neuronal development. Science (New
York, NY) 225:1271--1279 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/6474176}.

\hyperdef{}{ref-Hentschel1999Models}{}
Hentschel HGE, Ooyen A van (1999) Models of axon guidance and bundling
during development. Proceedings of the Royal Society of London B:
Biological Sciences 266:2231--2238 Available at:
\url{http://dx.doi.org/10.1098/rspb.1999.0913}.

\hyperdef{}{ref-Hines1997NEURON}{}
Hines ML, Carnevale NT (1997) The NEURON Simulation Environment. Neural
Computation 9:1179--1209 Available at:
\url{http://dx.doi.org/10.1162/neco.1997.9.6.1179}.

\hyperdef{}{ref-Hines2009NEURON}{}
Hines ML, Davison AP, Muller E (2009) NEURON and Python. Frontiers in
neuroinformatics 3 Available at:
\url{http://dx.doi.org/10.3389/neuro.11.001.2009}.

\hyperdef{}{ref-Holt1999Electrical}{}
Holt GR, Koch C (1999) Electrical Interactions via the Extracellular
Potential Near Cell Bodies. Journal of Computational Neuroscience
6:169--184 Available at:
\url{http://dx.doi.org/10.1023/a:1008832702585}.

\hyperdef{}{ref-kandel2000principles}{}
Kandel ER, Schwartz JH, Jessell TM, Others (2000) Principles of neural
science. McGraw-Hill New York.

\hyperdef{}{ref-Laughlin2010Oscillatory}{}
Laughlin MM, Verschooten E, Joris PX (2010) Oscillatory Dipoles As a
Source of Phase Shifts in Field Potentials in the Mammalian Auditory
Brainstem. The Journal of Neuroscience 30:13472--13487 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0294-10.2010}.

\hyperdef{}{ref-Linden2010Intrinsic}{}
Lindén H, Pettersen K, Einevoll G (2010) Intrinsic dendritic filtering
gives low-pass power spectra of local field potentials. Journal of
computational neuroscience 29:423--444 Available at:
\url{http://dx.doi.org/10.1007/s10827-010-0245-4}.

\hyperdef{}{ref-Linden2011Modeling}{}
Lindén H, Tetzlaff T, Potjans TC, Pettersen KH, Grün S, Diesmann M,
Einevoll GT (2011) Modeling the Spatial Reach of the LFP. Neuron
72:859--872 Available at:
\url{http://dx.doi.org/10.1016/j.neuron.2011.11.006}.

\hyperdef{}{ref-Nornes1972Temporal}{}
Nornes HO, Das GD (1972) Temporal pattern of neurogenesis in spinal
cord: cytoarchitecture and directed growth of axons. Proceedings of the
National Academy of Sciences of the United States of America
69:1962--1966 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/4114859}.

\hyperdef{}{ref-Nunez2006Electric}{}
Nunez PL, Srinivasan R (2006) Electric Fields of the Brain. Oxford
University Press. Available at:
\url{http://dx.doi.org/10.1093/acprof:oso/9780195050387.001.0001}.

\hyperdef{}{ref-Ray2011Different}{}
Ray S, Maunsell JHR (2011) Different Origins of Gamma Rhythm and
High-Gamma Activity in Macaque Visual Cortex. PLoS Biol 9:e1000610+
Available at: \url{http://dx.doi.org/10.1371/journal.pbio.1000610}.

\hyperdef{}{ref-Reimann2013Biophysically}{}
Reimann MW, Anastassiou CA, Perin R, Hill SL, Markram H, Koch C (2013) A
Biophysically Detailed Model of Neocortical Local Field Potentials
Predicts the Critical Role of Active Membrane Currents. Neuron
79:375--390 Available at:
\url{http://dx.doi.org/10.1016/j.neuron.2013.05.023}.

\hyperdef{}{ref-Schomburg2012Spiking}{}
Schomburg EW, Anastassiou CA, Buzsáki G, Koch C (2012) The Spiking
Component of Oscillatory Extracellular Potentials in the Rat
Hippocampus. The Journal of Neuroscience 32:11798--11811 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0656-12.2012}.

\hyperdef{}{ref-Schuz2002Human}{}
Schüz A, Braitenberg V (2002) The Human Cortical White Matter. In:
Cortical areas (Schüz A, ed), pp 377--386. Abingdon, UK: Taylor \&
Francis. Available at:
\url{http://dx.doi.org/10.4324/9780203219911/_chapter/_16}.

\hyperdef{}{ref-Schwarz1992Can}{}
Schwarz DW (1992) Can central neurons reproduce sound waveforms? An
analysis of the neurophonic potential in the laminar nucleus of the
chicken. The Journal of otolaryngology 21:30--38 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/1564747}.

\hyperdef{}{ref-Telenczuk2011Highfrequency}{}
Teleńczuk B, Baker SN, Herz AVM, Curio G (2011) High-frequency EEG
covaries with spike burst patterns detected in cortical neurons. Journal
of Neurophysiology 105:2951--2959 Available at:
\url{http://dx.doi.org/10.1152/jn.00327.2010}.

\hyperdef{}{ref-Telenczuk2015Correlates}{}
Teleńczuk B, Baker SN, Kempter R, Curio G (2015) Correlates of a single
cortical action potential in the epidural EEG. NeuroImage 109:357--367
Available at: \url{http://dx.doi.org/10.1016/j.neuroimage.2014.12.057}.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[a4paper]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations},
            pdfauthor={Thomas McColgan, Hermann Wagner, Catherine E Carr, Richard Kempter},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing

\title{Extracellular potentials generated by axonal projections are shaped by
patterns of bifurcations and terminations}
\author{Thomas McColgan, Hermann Wagner, Catherine E Carr, Richard Kempter}
\date{\today}

@@

Extracellular field potentials (EFPs) are at the heart of many
experimental methods used to examine the inner workings of the brain.
This includes invasive methods such as Local Field Potentials (LFP),
Current Source Density (CSD) and Multiunit Activity (MUA) as well as
noninvasive methods such as Electroencephalogram (EEG),
Electrocorticogram (ECoG) and the auditory brainstem response (ABR)
(Nunez and Srinivasan, 2006; Brette and Destexhe, 2012). The origins of
these EFPs, especially in cases in which the activity is not clearly
attributable to a single cell, is a matter of debate.

EFPs in the brain were long thought to be primarily synaptic in origin
(Buzsáki et al., 2012). As a consequence, many modeling studies focussed
on the extracellular fields induced by synaptic currents on the
dendrites and the soma of a postsynaptic neuron (Holt and Koch, 1999;
Gold et al., 2006; Lindén et al., 2010, 2011; Einevoll et al., 2013;
Fernández-Ruiz et al., 2013). However, a number of recent data analysis
and modeling efforts have revealed that active, non-synaptic membrane
currents can play an important role in generating EFPs (Ray and
Maunsell, 2011; Belluscio et al., 2012; Schomburg et al., 2012; Reimann
et al., 2013; Anastassiou et al., 2015), including far reaching
potentials detectable at the scalp (Teleńczuk et al., 2011, 2015).
However, currents from the axon are still thought to be so small as to
be of minor importance for the EFP.

One of the reasons for the traditional assumption that axonal currents
contribute little to the EFP is that the far field of an action
potential traveling along an idealized straight axon is quadrupolar,
meaning that its far field decays faster with distance than synaptic
sources, which are thought to be dipolar (Nunez and Srinivasan, 2006).
However, in experimental studies that recorded the EFP of axonal
responses, the EFP often has a dipolar structure. For example, Blot and
Barbour (2014) report an EFP with the characteristic dipolar structure
recorded from the vicinity of cerebellar Purkinje cell axons.

\begin{itemize}
\tightlist
\item
  find more examples in literature
\end{itemize}

In the results section we will present a further new example of such a
field potential from the auditory brainstem of the barn owl.

This discrepancy between theory and experiment raises the open question
of how axons are able to generate dipolar field potentials.

The aim of this study is to show how dipolar far fields in the EFP of
axons can be explained by the axons' anatomical structure. In
particular, we show that the branchings and termonations of axons in
their projection area lead to a dipolar EFP structure. Such axon
bundles, often also called nerves or fascicles, exist throughout the
peripheral and central nervous system of most animals (Nornes and Das,
1972; Goodman et al., 1984; Hentschel and Ooyen, 1999; Kandel et al.,
2000). The white matter of the mammalian brain can be viewed as an
agglomeration of such bundles (Schüz and Braitenberg, 2002). We
therefore predict pronounced contributions of nerves or fascicles to
EFPs, which are neglected in all current models.

Properties of the EFP due to axonal projections are first investigated
by a numerical model based on forward simulation (Holt and Koch, 1999;
Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we outline the mechanisms underlying the observed effects by
means of a second, analytically tractable, model of a generic axon
bundle.

@@

\emph{Axonal projections generate a dipole-like field potential.} To
understand how axons can generate a dipolar EFP, we simulated a parallel
fiber bundle that at first runs at constant diameter without
bifurcations and then reaches a projection zone. Within this projection
zone, the fibers first bifurcate, which increases the number of fibers.
Finally, the axons reach a termination zone, in which the number of
fibers decreases to zero (Figure 1A).

The spiking activity of a generic axon bundle was simulated by
stimulating the fibers with a background spontaneous firing rate of 100
spikes/second, upon which a pulse of increased activity was added. This
pulse of activity could relate to any kind of evoked activity in the
nervous system, such as a sensory stimulation, motor activity or a
spontaneous transient increase in population spiking activity. We chose
a gaussian pulse with an amplitude of 2900 spikes/second and a standard
deviation of 2.8 ms.

To characterize the spatiotemporal dynamics of the evoked EFP, the
potential was calculated for several locations along the axon trunk.
Because we first want to study the effect of the rate pulse on the EFP,
we low-pass filtered the EFP responses with a cut-off frequency of 1 kHz
(Figure 1B). The distribution of maximum amplitudes of these responses
are shown by the colored contour lines in Figure 1A and the colored
traces in Figure 1B. Surrounding the projection zone of the axon bundle,
amplitudes show the double-lobed shape typical of a dipole.

In Figure 1B, all the responses showed a biphasic deflection that was
elicited by the population firing rate pulse. This deflection was
visible at all recording locations. The biphasic deflection reverses
polarity in the middle of the bifurcation zone of the bundle (red box in
Figure 1B). The polarity reversal occurs by a decrease of the amplitude
to zero and subsequent reappearance with reversed polarity, as opposed
to a polarity reversal through a gradual shift in phase. This behaviour
is also typical for a dipolar field potential.

The point of the polarity reversal coincides roughly with the middle of
the projection zone. This means that the absolute value of the response
amplitude reaches a local minimum just when the number of axonal fibers
reaches a maximum. To better understand how anatomical features and
response amplitude are related, we compared the signed maximum value
(meaning the \emph{signed} value corresponding to the maximum
\emph{absolute} amplitude) of the EFP with the difference in the density
of branchings and terminations (Figure 1C). Interestingly, the amplitude
of the maximum response (red trace) follows the distribution of
branchings and terminations (black histogram).

@@

\hypertarget{refs}{}
\hypertarget{ref-Anastassiou2015Cell}{}
Anastassiou CA, Perin R, Buzsáki G, Markram H, Koch C (2015) Cell type-
and activity-dependent extracellular correlates of intracellular
spiking. Journal of Neurophysiology 114:608--623 Available at:
\url{http://dx.doi.org/10.1152/jn.00628.2014}.

\hypertarget{ref-Belluscio2012CrossFrequency}{}
Belluscio MA, Mizuseki K, Schmidt R, Kempter R, Buzsáki G (2012)
Cross-Frequency Phase--Phase Coupling between Theta and Gamma
Oscillations in the Hippocampus. The Journal of Neuroscience 32:423--435
Available at: \url{http://dx.doi.org/10.1523/jneurosci.4122-11.2012}.

\hypertarget{ref-Blot2014Ultrarapid}{}
Blot A, Barbour B (2014) Ultra-rapid axon-axon ephaptic inhibition of
cerebellar Purkinje cells by the pinceau. Nat Neurosci 17:289--295
Available at: \url{http://dx.doi.org/10.1038/nn.3624}.

\hypertarget{ref-Brette2012Handbook}{}
Brette R, Destexhe A eds. (2012) Handbook of Neural Activity
Measurement. Cambridge: Cambridge University Press. Available at:
\url{http://dx.doi.org/10.1017/cbo9780511979958}.

\hypertarget{ref-Buzsaki2012Origin}{}
Buzsáki G, Anastassiou CA, Koch C (2012) The origin of extracellular
fields and currents--EEG, ECoG, LFP and spikes. Nature reviews
Neuroscience 13:407--420 Available at:
\url{http://dx.doi.org/10.1038/nrn3241}.

\hypertarget{ref-carr90}{}
Carr CE, Konishi M (1990) A circuit for detection of interaural time
differences in the brain stem of the barn owl. The Journal of
Neuroscience 10:3227--3246 Available at:
\url{http://www.jneurosci.org/content/10/10/3227.abstract}.

\hypertarget{ref-Einevoll2013Modelling}{}
Einevoll GT, Kayser C, Logothetis NK, Panzeri S (2013) Modelling and
analysis of local field potentials for studying the function of cortical
circuits. Nature reviews Neuroscience 14:770--785 Available at:
\url{http://dx.doi.org/10.1038/nrn3599}.

\hypertarget{ref-FernandezRuiz2013Cytoarchitectonic}{}
Fernández-Ruiz A, Muñoz S, Sancho M, Makarova J, Makarov VA, Herreras O
(2013) Cytoarchitectonic and Dynamic Origins of Giant Positive Local
Field Potentials in the Dentate Gyrus. The Journal of Neuroscience
33:15518--15532 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0338-13.2013}.

\hypertarget{ref-Gold2006Origin}{}
Gold C, Henze DA, Koch C, Buzsáki G (2006) On the origin of the
extracellular action potential waveform: A modeling study. Journal of
neurophysiology 95:3113--3128 Available at:
\url{http://dx.doi.org/10.1152/jn.00979.2005}.

\hypertarget{ref-Goldwyn2014Model}{}
Goldwyn JH, Laughlin MM, Verschooten E, Joris PX, Rinzel J (2014) A
Model of the Medial Superior Olive Explains Spatiotemporal Features of
Local Field Potentials. The Journal of Neuroscience 34:11705--11722
Available at: \url{http://dx.doi.org/10.1523/jneurosci.0175-14.2014}.

\hypertarget{ref-Goodman1984Cell}{}
Goodman CS, Bastiani MJ, Doe CQ, Lac S du, Helfand SL, Kuwada JY, Thomas
JB (1984) Cell recognition during neuronal development. Science (New
York, NY) 225:1271--1279 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/6474176}.

\hypertarget{ref-Hentschel1999Models}{}
Hentschel HGE, Ooyen A van (1999) Models of axon guidance and bundling
during development. Proceedings of the Royal Society of London B:
Biological Sciences 266:2231--2238 Available at:
\url{http://dx.doi.org/10.1098/rspb.1999.0913}.

\hypertarget{ref-Hines1997NEURON}{}
Hines ML, Carnevale NT (1997) The NEURON Simulation Environment. Neural
Computation 9:1179--1209 Available at:
\url{http://dx.doi.org/10.1162/neco.1997.9.6.1179}.

\hypertarget{ref-Hines2009NEURON}{}
Hines ML, Davison AP, Muller E (2009) NEURON and Python. Frontiers in
neuroinformatics 3 Available at:
\url{http://dx.doi.org/10.3389/neuro.11.001.2009}.

\hypertarget{ref-Holt1999Electrical}{}
Holt GR, Koch C (1999) Electrical Interactions via the Extracellular
Potential Near Cell Bodies. Journal of Computational Neuroscience
6:169--184 Available at:
\url{http://dx.doi.org/10.1023/a:1008832702585}.

\hypertarget{ref-kandel2000principles}{}
Kandel ER, Schwartz JH, Jessell TM, Others (2000) Principles of neural
science. McGraw-Hill New York.

\hypertarget{ref-Laughlin2010Oscillatory}{}
Laughlin MM, Verschooten E, Joris PX (2010) Oscillatory Dipoles As a
Source of Phase Shifts in Field Potentials in the Mammalian Auditory
Brainstem. The Journal of Neuroscience 30:13472--13487 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0294-10.2010}.

\hypertarget{ref-Linden2010Intrinsic}{}
Lindén H, Pettersen K, Einevoll G (2010) Intrinsic dendritic filtering
gives low-pass power spectra of local field potentials. Journal of
computational neuroscience 29:423--444 Available at:
\url{http://dx.doi.org/10.1007/s10827-010-0245-4}.

\hypertarget{ref-Linden2011Modeling}{}
Lindén H, Tetzlaff T, Potjans TC, Pettersen KH, Grün S, Diesmann M,
Einevoll GT (2011) Modeling the Spatial Reach of the LFP. Neuron
72:859--872 Available at:
\url{http://dx.doi.org/10.1016/j.neuron.2011.11.006}.

\hypertarget{ref-Nornes1972Temporal}{}
Nornes HO, Das GD (1972) Temporal pattern of neurogenesis in spinal
cord: cytoarchitecture and directed growth of axons. Proceedings of the
National Academy of Sciences of the United States of America
69:1962--1966 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/4114859}.

\hypertarget{ref-Nunez2006Electric}{}
Nunez PL, Srinivasan R (2006) Electric Fields of the Brain. Oxford
University Press. Available at:
\url{http://dx.doi.org/10.1093/acprof:oso/9780195050387.001.0001}.

\hypertarget{ref-Ray2011Different}{}
Ray S, Maunsell JHR (2011) Different Origins of Gamma Rhythm and
High-Gamma Activity in Macaque Visual Cortex. PLoS Biol 9:e1000610+
Available at: \url{http://dx.doi.org/10.1371/journal.pbio.1000610}.

\hypertarget{ref-Reimann2013Biophysically}{}
Reimann MW, Anastassiou CA, Perin R, Hill SL, Markram H, Koch C (2013) A
Biophysically Detailed Model of Neocortical Local Field Potentials
Predicts the Critical Role of Active Membrane Currents. Neuron
79:375--390 Available at:
\url{http://dx.doi.org/10.1016/j.neuron.2013.05.023}.

\hypertarget{ref-Schomburg2012Spiking}{}
Schomburg EW, Anastassiou CA, Buzsáki G, Koch C (2012) The Spiking
Component of Oscillatory Extracellular Potentials in the Rat
Hippocampus. The Journal of Neuroscience 32:11798--11811 Available at:
\url{http://dx.doi.org/10.1523/jneurosci.0656-12.2012}.

\hypertarget{ref-Schuz2002Human}{}
Schüz A, Braitenberg V (2002) The Human Cortical White Matter. In:
Cortical areas (Schüz A, ed), pp 377--386. Abingdon, UK: Taylor \&
Francis. Available at:
\url{http://dx.doi.org/10.4324/9780203219911/_chapter/_16}.

\hypertarget{ref-Schwarz1992Can}{}
Schwarz DW (1992) Can central neurons reproduce sound waveforms? An
analysis of the neurophonic potential in the laminar nucleus of the
chicken. The Journal of otolaryngology 21:30--38 Available at:
\url{http://view.ncbi.nlm.nih.gov/pubmed/1564747}.

\hypertarget{ref-Telenczuk2011Highfrequency}{}
Teleńczuk B, Baker SN, Herz AVM, Curio G (2011) High-frequency EEG
covaries with spike burst patterns detected in cortical neurons. Journal
of Neurophysiology 105:2951--2959 Available at:
\url{http://dx.doi.org/10.1152/jn.00327.2010}.

\hypertarget{ref-Telenczuk2015Correlates}{}
Teleńczuk B, Baker SN, Kempter R, Curio G (2015) Correlates of a single
cortical action potential in the epidural EEG. NeuroImage 109:357--367
Available at: \url{http://dx.doi.org/10.1016/j.neuroimage.2014.12.057}.
